---
title: "Data organization"
author: "Alper Cevirgel"
output: pdf_document
---

# library
```{r message=FALSE, warning=FALSE, include=FALSE}
source("codes/libraries_functions.R") #load the libraries
analysis.path <- getwd()
```

# HI data organization
## Import HI data
```{r}
## import data, contains collection day, assay type, sample type
HI_all <- read_excel("/Users/alpercevirgel/Documents/Alper RIVM/1_Complete_analysis_per_project/VITAL database/VITAL database Alper.xlsx",
                           sheet = "HI_influenza",
                           skip = 1) %>% 
  mutate_at(c("subject_identifier"), list(~as.character(.))) %>% 
  as.data.frame()

```

# import additional datasets, demographics, trucount, CMV, immunotype
## Subject information, age, age group, sex
```{r}
## subset the age, age groups and sex data 
vital.subject <- read_excel("/Users/alpercevirgel/Documents/Alper RIVM/1_Complete_analysis_per_project/VITAL database/VITAL database Alper.xlsx",
                            sheet = "subject",
                            skip = 1) %>% 
  mutate_at(c("sex","Age_group_T0"), list(~factor(.))) %>%
  mutate_at(c("subject_identifier"), list(~as.character(.))) %>%
  dplyr::rename(age_group=Age_group_T0,
                age=Age_T0,
                BMI=BMI_T0) %>%
  dplyr::select(subject_identifier,sex,age,BMI,age_group) %>%
  as.data.frame()
head(vital.subject) 
```

## Chronic viral infection data
```{r}
## add chronic herpes virus infection data, and create factors for CMV+/- & EBV+/- combinations
chronic_viral_inf <- read_excel("/Users/alpercevirgel/Documents/Alper RIVM/1_Complete_analysis_per_project/VITAL database/VITAL database Alper.xlsx", 
                                       sheet = "CMV_EBV_VZV",
                                       skip = 1,na = "n/a") %>%
  dplyr::select(subject_identifier,MIA_CMV_Titer,MIA_CMV_Seropositivity,MIA_EBV_Titer,MIA_EBV_Seropositivity,MIA_VZV_Titer,MIA_VZV_Seropositivity) %>%
  mutate(CMV_EBV_combined = case_when(
      MIA_CMV_Seropositivity == "1" & MIA_EBV_Seropositivity == "1"  ~ "CMV+EBV+",
      MIA_CMV_Seropositivity == "2" & MIA_EBV_Seropositivity == "2"  ~ "CMV-EBV-",
      MIA_CMV_Seropositivity == "3" | MIA_EBV_Seropositivity == "3"  ~ "Borderline",
      MIA_CMV_Seropositivity == "1" & MIA_EBV_Seropositivity == "2"  ~ "CMV+EBV-",
      MIA_CMV_Seropositivity == "2" & MIA_EBV_Seropositivity == "1"  ~ "CMV-EBV+",
),.after="subject_identifier") %>%
  mutate_at(c("subject_identifier"), list(~as.character(.))) %>%
  mutate_at(c("MIA_CMV_Seropositivity","MIA_EBV_Seropositivity","MIA_VZV_Seropositivity","CMV_EBV_combined"), list(~factor(.)))

head(chronic_viral_inf)
```



## Trucount data (IQV)
```{r}
## import data, contains the influenza vaccination trucount data, age, timepoint 
## Import raw data, 
vital.truc_influ <- read_excel("/Users/alpercevirgel/Documents/Alper RIVM/1_Complete_analysis_per_project/VITAL database/VITAL database Alper.xlsx",
                               sheet = "trucount_influenza_vac",
                               skip = 1) %>% 
  mutate_at(c("sex","age_group","timepoint","timepoint_days"), list(~factor(.))) %>%
  mutate_at(c("subject_identifier"), list(~as.character(.))) %>%
  as.data.frame()

## some additional variables added later
influ_truc_additional <- read_excel("/Users/alpercevirgel/Documents/Alper RIVM/1_Complete_analysis_per_project/VITAL database/VITAL database Alper.xlsx",
                                    sheet = "extented_influ_trucount_subsets",
                                    skip = 1) %>% 
  dplyr::select(-c(timepoint, subject_identifier))%>%
  mutate_at(c("sample_identifier"), list(~as.character(.)))

## merge
vital.truc_influ_additional <- vital.truc_influ %>%
  left_join(influ_truc_additional, by="sample_identifier") %>%
  dplyr::select(-c(fcs_file,age_group,age,sex,timepoint_days,sample_identifier)) 


## filter timepoints
vital.truc_influ_d0_additional <- vital.truc_influ_additional %>% filter(timepoint %in% "A")
vital.truc_influ_d2_additional <- vital.truc_influ_additional %>% filter(timepoint %in% "B")
vital.truc_influ_d7_additional <- vital.truc_influ_additional %>% filter(timepoint %in% "C")

## add timepoint info to variables 
### day 0
vital.truc_influ_d0_additional <- rename_with(vital.truc_influ_d0_additional, ~(gsub("_per", "_D0_per", .x, fixed = TRUE)))
vital.truc_influ_d0_additional <- rename_with(vital.truc_influ_d0_additional, ~(gsub("_num", "_D0_num", .x, fixed = TRUE)))

vital.truc_influ_d0_additional <- vital.truc_influ_d0_additional %>%
  rename(CD4TnTm_D0=CD4TnTm,
         CD8TnTm_D0=CD8TnTm)

### day 1-2
vital.truc_influ_d2_additional <- rename_with(vital.truc_influ_d2_additional, ~(gsub("_per", "_D2_per", .x, fixed = TRUE)))
vital.truc_influ_d2_additional <- rename_with(vital.truc_influ_d2_additional, ~(gsub("_num", "_D2_num", .x, fixed = TRUE)))

vital.truc_influ_d2_additional <- vital.truc_influ_d2_additional %>%
  rename(CD4TnTm_D2=CD4TnTm,
         CD8TnTm_D2=CD8TnTm)

### day 7
vital.truc_influ_d7_additional <- rename_with(vital.truc_influ_d7_additional, ~(gsub("_per", "_D7_per", .x, fixed = TRUE)))
vital.truc_influ_d7_additional <- rename_with(vital.truc_influ_d7_additional, ~(gsub("_num", "_D7_num", .x, fixed = TRUE)))

vital.truc_influ_d7_additional <- vital.truc_influ_d7_additional %>%
  rename(CD4TnTm_D7=CD4TnTm,
         CD8TnTm_D7=CD8TnTm)
```

# Calculate immune subset kinetics between Day0-Day1/2 and Day0-Day7
## Kinetics Day0 between Day1-2
```{r}
#### Day 1-2 vs Day0 ####
df_kinetics_D0D12 <- vital.truc_influ_additional %>% filter(timepoint %in% c("A","B") )

## identifying subject IDs with both Day0 and Day1-2 data points
subject_ID_D0D12 <- df_kinetics_D0D12[duplicated(df_kinetics_D0D12$subject_identifier),]$subject_identifier

## removing the subject IDs that do not contain both D0D12 data , D12 means Day1-Day2
print("dims before non-D0D12 datapoints removed") ; dim(df_kinetics_D0D12)
df_kinetics_D0D12 <- df_kinetics_D0D12[(df_kinetics_D0D12$subject_identifier %in% subject_ID_D0D12),]
print("dims after non-D0D12 datapoints removed") ; dim(df_kinetics_D0D12)

## subset the day 0 and day 1-2 data
df_kinetics_D0 <- df_kinetics_D0D12 %>% filter(timepoint %in% c("A"))
df_kinetics_D12 <- df_kinetics_D0D12 %>% filter(timepoint %in% c("B"))

## check if the IDs match
print("Do the subject IDs match?");all.equal(df_kinetics_D0$subject_identifier,df_kinetics_D12$subject_identifier)


## calculate the log2 fold change
df_kinetics_D12_D0 <- as.data.frame(c()) ## set as a dataframe
df_kinetics_D12_D0 <- df_kinetics_D0[c(1:2)] ## add participant data
df_kinetics_D12_D0_log2 <- as.data.frame(c()) ## set as a dataframe
df_kinetics_D12_D0_log2 <- df_kinetics_D0[c(1:2)] ## add participant data
NCOL <- ncol(df_kinetics_D0)

df_kinetics_D12_D0[c(3:NCOL)] <- ((df_kinetics_D12[c(3:NCOL)])/(df_kinetics_D0[c(3:NCOL)])) ## calculate the fold changes
truc_d12d0_raw <- df_kinetics_D12_D0 %>% dplyr::select(-timepoint)

df_kinetics_D12_D0_log2[c(3:NCOL)] <- log2((df_kinetics_D12[c(3:NCOL)])/(df_kinetics_D0[c(3:NCOL)])) ## calculate the fold changes
df_kinetics_D12_D0_influ_log2fc <- df_kinetics_D12_D0_log2 %>% dplyr::select(-c(timepoint)) ##remove non-necessary columns 

## rename the columns
df_kinetics_D12_D0_influ_log2fc <- rename_with(df_kinetics_D12_D0_influ_log2fc, ~(gsub("_per", "_D12D0_per", .x, fixed = TRUE)))
df_kinetics_D12_D0_influ_log2fc <- rename_with(df_kinetics_D12_D0_influ_log2fc, ~(gsub("_num", "_D12D0_num", .x, fixed = TRUE)))
df_kinetics_D12_D0_influ_log2fc <- df_kinetics_D12_D0_influ_log2fc %>%
  rename(CD4TnTm_D12D0=CD4TnTm,
         CD8TnTm_D12D0=CD8TnTm)

truc_d12d0_raw <- rename_with(truc_d12d0_raw, ~(gsub("_per", "_D12D0_per", .x, fixed = TRUE)))
truc_d12d0_raw <- rename_with(truc_d12d0_raw, ~(gsub("_num", "_D12D0_num", .x, fixed = TRUE)))
truc_d12d0_raw <- truc_d12d0_raw %>%
  rename(CD4TnTm_D12D0=CD4TnTm,
         CD8TnTm_D12D0=CD8TnTm)

## calculate the log2 fold change (adding 0.01 to every cell)
df_kinetics_D12_D0 <- as.data.frame(c()) ## set as a dataframe
df_kinetics_D12_D0 <- df_kinetics_D0[c(1:2)] ## add participant data
df_kinetics_D12_D0_log2 <- as.data.frame(c()) ## set as a dataframe
df_kinetics_D12_D0_log2 <- df_kinetics_D0[c(1:2)] ## add participant data
NCOL <- ncol(df_kinetics_D0)


df_kinetics_D12_D0[c(3:NCOL)] <- ((df_kinetics_D12[c(3:NCOL)]+0.01)/(df_kinetics_D0[c(3:NCOL)]+0.01)) ## calculate the fold changes
truc_d12d0_raw_p01 <- df_kinetics_D12_D0 %>% dplyr::select(-timepoint)

df_kinetics_D12_D0_log2[c(3:NCOL)] <- log2((df_kinetics_D12[c(3:NCOL)]+0.01)/(df_kinetics_D0[c(3:NCOL)]+0.01)) ## calculate the fold changes
df_kinetics_D12_D0_influ_log2fc_p01 <- df_kinetics_D12_D0_log2 %>% dplyr::select(-c(timepoint)) ##remove non-necessary columns 

## rename the columns
df_kinetics_D12_D0_influ_log2fc_p01 <- rename_with(df_kinetics_D12_D0_influ_log2fc_p01, ~(gsub("_per", "_D12D0_per", .x, fixed = TRUE)))
df_kinetics_D12_D0_influ_log2fc_p01 <- rename_with(df_kinetics_D12_D0_influ_log2fc_p01, ~(gsub("_num", "_D12D0_num", .x, fixed = TRUE)))
df_kinetics_D12_D0_influ_log2fc_p01 <- df_kinetics_D12_D0_influ_log2fc_p01 %>%
  rename(CD4TnTm_D12D0=CD4TnTm,
         CD8TnTm_D12D0=CD8TnTm)

truc_d12d0_raw_p01 <- rename_with(truc_d12d0_raw_p01, ~(gsub("_per", "_D12D0_per", .x, fixed = TRUE)))
truc_d12d0_raw_p01 <- rename_with(truc_d12d0_raw_p01, ~(gsub("_num", "_D12D0_num", .x, fixed = TRUE)))
truc_d12d0_raw_p01 <- truc_d12d0_raw_p01 %>%
  rename(CD4TnTm_D12D0=CD4TnTm,
         CD8TnTm_D12D0=CD8TnTm)
```

## Kinetics Day0 between Day7
```{r}
#### Day 7 vs Day0 ####
df_kinetics_D0D7 <- vital.truc_influ_additional %>% filter(timepoint %in% c("A","C") )

## identifying subject IDs with both Day0 and Day7 data points
subject_ID_D0D7 <- df_kinetics_D0D7[duplicated(df_kinetics_D0D7$subject_identifier),]$subject_identifier

## removing the subject IDs that do not contain both D0D7 data
print("dims before non-D0D7 datapoints removed") ; dim(df_kinetics_D0D7)
df_kinetics_D0D7 <- df_kinetics_D0D7[(df_kinetics_D0D7$subject_identifier %in% subject_ID_D0D7),]
print("dims after non-D0D7 datapoints removed") ; dim(df_kinetics_D0D7)

## subset the day 0 and day 7 data
df_kinetics_D0 <- df_kinetics_D0D7 %>% filter(timepoint %in% c("A"))
df_kinetics_D7 <- df_kinetics_D0D7 %>% filter(timepoint %in% c("C"))

## check if the IDs match
print("Do the subject IDs match?");all.equal(df_kinetics_D0$subject_identifier,df_kinetics_D7$subject_identifier)

## calculate the log2 fold change
df_kinetics_D7_D0 <- as.data.frame(c()) ## set as a dataframe
df_kinetics_D7_D0 <- df_kinetics_D0[c(1,2)] ## add participant data
df_kinetics_D7_D0_log2 <- as.data.frame(c()) ## set as a dataframe
df_kinetics_D7_D0_log2 <- df_kinetics_D0[c(1,2)] ## add participant data
NCOL <- ncol(df_kinetics_D0)

df_kinetics_D7_D0[c(3:NCOL)] <- ((df_kinetics_D7[c(3:NCOL)])/(df_kinetics_D0[c(3:NCOL)])) ## calculate the fold changes
truc_d7d0_raw <- df_kinetics_D7_D0 %>% dplyr::select(-timepoint)

df_kinetics_D7_D0_log2[c(3:NCOL)] <- log2((df_kinetics_D7[c(3:NCOL)])/(df_kinetics_D0[c(3:NCOL)])) ## calculate the fold changes
df_kinetics_D7_D0_influ_log2fc <- df_kinetics_D7_D0_log2 %>% dplyr::select(-c(timepoint)) ##remove non-necessary columns 

## rename the columns
df_kinetics_D7_D0_influ_log2fc <- rename_with(df_kinetics_D7_D0_influ_log2fc, ~(gsub("_per", "_D7D0_per", .x, fixed = TRUE)))
df_kinetics_D7_D0_influ_log2fc <- rename_with(df_kinetics_D7_D0_influ_log2fc, ~(gsub("_num", "_D7D0_num", .x, fixed = TRUE)))
df_kinetics_D7_D0_influ_log2fc <- df_kinetics_D7_D0_influ_log2fc %>%
  rename(CD4TnTm_D7D0=CD4TnTm,
         CD8TnTm_D7D0=CD8TnTm)

truc_d7d0_raw <- rename_with(truc_d7d0_raw, ~(gsub("_per", "_D7D0_per", .x, fixed = TRUE)))
truc_d7d0_raw <- rename_with(truc_d7d0_raw, ~(gsub("_num", "_D7D0_num", .x, fixed = TRUE)))
truc_d7d0_raw <- truc_d7d0_raw %>%
  rename(CD4TnTm_D7D0=CD4TnTm,
         CD8TnTm_D7D0=CD8TnTm)

## calculate the log2 fold change (adding 0.01 to every cell)
df_kinetics_D7_D0 <- as.data.frame(c()) ## set as a dataframe
df_kinetics_D7_D0 <- df_kinetics_D0[c(1,2)] ## add participant data
df_kinetics_D7_D0_log2 <- as.data.frame(c()) ## set as a dataframe
df_kinetics_D7_D0_log2 <- df_kinetics_D0[c(1,2)] ## add participant data
NCOL <- ncol(df_kinetics_D0)

df_kinetics_D7_D0[c(3:NCOL)] <- ((df_kinetics_D7[c(3:NCOL)]+0.01)/(df_kinetics_D0[c(3:NCOL)]+0.01)) ## calculate the fold changes
truc_d7d0_raw_p01 <- df_kinetics_D7_D0 %>% dplyr::select(-timepoint)

df_kinetics_D7_D0_log2[c(3:NCOL)] <- log2((df_kinetics_D7[c(3:NCOL)]+0.01)/(df_kinetics_D0[c(3:NCOL)]+0.01)) ## calculate the fold changes
df_kinetics_D7_D0_influ_log2fc_p01 <- df_kinetics_D7_D0_log2 %>% dplyr::select(-c(timepoint)) ##remove non-necessary columns 

## rename the columns
df_kinetics_D7_D0_influ_log2fc_p01 <- rename_with(df_kinetics_D7_D0_influ_log2fc_p01, ~(gsub("_per", "_D7D0_per", .x, fixed = TRUE)))
df_kinetics_D7_D0_influ_log2fc_p01 <- rename_with(df_kinetics_D7_D0_influ_log2fc_p01, ~(gsub("_num", "_D7D0_num", .x, fixed = TRUE)))
df_kinetics_D7_D0_influ_log2fc_p01 <- df_kinetics_D7_D0_influ_log2fc_p01 %>%
  rename(CD4TnTm_D7D0=CD4TnTm,
         CD8TnTm_D7D0=CD8TnTm)

truc_d7d0_raw_p01 <- rename_with(truc_d7d0_raw_p01, ~(gsub("_per", "_D7D0_per", .x, fixed = TRUE)))
truc_d7d0_raw_p01 <- rename_with(truc_d7d0_raw_p01, ~(gsub("_num", "_D7D0_num", .x, fixed = TRUE)))
truc_d7d0_raw_p01 <- truc_d7d0_raw_p01 %>%
  rename(CD4TnTm_D7D0=CD4TnTm,
         CD8TnTm_D7D0=CD8TnTm)

```

# HAI GMT Day 0 levels
## Create GMT T0 factor groups
```{r}
HI_df <- HI_all %>%
  mutate(T0_l3= ifelse(GMT_T0<40, "<40",
                       ifelse(GMT_T0 <100 & GMT_T0>=40 , "40-80",
                              ifelse(GMT_T0 >= 100, ">80", NA)))) %>%
  mutate_at(c("T0_l3"), list(~factor(.)))

HI_df$T0_l3 <- factor(HI_df$T0_l3, levels = c("<40","40-80",">80"))

table(HI_df$T0_l3)

HI_df_subject <- HI_df %>%
  left_join(vital.subject,by="subject_identifier")
```

## Immunotype
```{r}
## import immunotypes
immunotype <- readRDS(
  "/Users/alpercevirgel/Documents/Alper RIVM/4_Manuscripts/Chapter_1/R_analysis/vital.truc_influ_d0_cluster.rds") %>%
  dplyr::select(subject_identifier,cluster_number)

## merge with antibody data
HI_immunotype <- HI_df %>% 
  left_join(immunotype, by="subject_identifier") %>% 
  drop_na(sero_res3, cluster_number)

## reorder responder levels
HI_immunotype$sero_res3 <- factor(HI_immunotype$sero_res3, levels = c("non_responder","low_responder","high_responder"))

```

## add CMV and cluster data
```{r}
HI_df_subject_CVI <- HI_df_subject %>% left_join(chronic_viral_inf, by="subject_identifier") %>%
    mutate_at(c("sero_res2","sero_res3"), list(~factor(.)))

head(HI_df_subject_CVI)

```

## stat analysis GMT T0 factors & age
```{r}
## melt
df_stat_melt <- HI_df_subject %>% 
  dplyr::select(T0_l3,age) %>%
  reshape2::melt(.)

### kruskal.test and Dunn's test
## With kruskal.test we identify overall which subsets are significant and then Dunn's test tells us which are different between pairs of clusters
kw.test.all <- df_stat_melt %>% 
  group_by(variable) %>% 
  kruskal_test(value ~ T0_l3) %>% 
  adjust_pvalue(method = "BH") 
kw.test.all.eff.size <- df_stat_melt %>% 
  group_by(variable) %>% 
  kruskal_effsize(value ~ T0_l3) 
combine.kw <- kw.test.all %>% 
  left_join(kw.test.all.eff.size, by="variable") %>% 
  filter(p.adj < 0.05)

## Number of variables with p.adj-value <0.05
length(combine.kw$variable)

## table summarizing effect size of the significant variables 
table(combine.kw$magnitude)

### Post-hoc analysis dunn's test, pairwise multiple comparisons to check which are significant in pairwise analysis.
dun.test.pos <- df_stat_melt %>% 
  group_by(variable) %>% 
  dunn_test(value ~ T0_l3, p.adjust.method = "BH", detailed = TRUE)
dun.test.pos <- dun.test.pos %>% 
  filter(p.adj.signif != "ns") # %>% 
  #rename(cluster_number_1 = group1) %>% 
  #rename(cluster_number_2 = group2)
head(dun.test.pos)

unique(dun.test.pos$variable)
```

## stat analysis GMT T0 levels & immunotype
```{r}
immunotype %>% 
  left_join(HI_df, by="subject_identifier") %>%
  drop_na(cluster_number,GMT_T4) %>%
  dplyr::select(cluster_number,GMT_T4) %>%
  droplevels() %>%
  group_by(cluster_number) %>%
  summarise(mean(GMT_T4)) %>% 
  arrange(`mean(GMT_T4)`) 

immunotype %>% 
  left_join(HI_df, by="subject_identifier") %>%
  drop_na(cluster_number,GMT_T4) %>%
  dplyr::select(cluster_number,GMT_T4) %>%
  droplevels() %>%
  group_by(cluster_number) %>%
  summarise(median(GMT_T4)) %>% 
  arrange(`median(GMT_T4)`) 


immunotype %>% 
  left_join(HI_df, by="subject_identifier") %>%
  drop_na(cluster_number,GMT_T4T0) %>%
  dplyr::select(cluster_number,GMT_T4T0) %>%
  droplevels() %>%
  group_by(cluster_number) %>%
  summarise(mean(GMT_T4T0)) %>% 
  arrange(`mean(GMT_T4T0)`) 

immunotype %>% 
  left_join(HI_df, by="subject_identifier") %>%
  drop_na(cluster_number,GMT_T4T0) %>%
  dplyr::select(cluster_number,GMT_T4T0) %>%
  droplevels() %>%
  group_by(cluster_number) %>%
  summarise(median(GMT_T4T0)) %>% 
  arrange(`median(GMT_T4T0)`) 
```

### D0
```{r}
## select the dataset
df_stat <- immunotype %>% 
  left_join(HI_df, by="subject_identifier") %>%
  drop_na(cluster_number) %>%
  dplyr::select(subject_identifier,cluster_number,GMT_T0) %>%
  droplevels()

## subject identifier and cluster number as.character
df_stat %<>% mutate_at(
  c("subject_identifier","cluster_number"),
  list(~as.character(.)))

## melt
df_stat_melt <- df_stat %>% 
  reshape2::melt(.)

### kruskal.test and Dunn's test
## With kruskal.test we identify overall which subsets are significant and then Dunn's test tells us which are different between pairs of clusters
kw.test.all <- df_stat_melt %>% 
  group_by(variable) %>% 
  kruskal_test(value ~ cluster_number) %>% 
  adjust_pvalue(method = "BH") 
kw.test.all.eff.size <- df_stat_melt %>% 
  group_by(variable) %>% 
  kruskal_effsize(value ~ cluster_number) 
combine.kw <- kw.test.all %>% 
  left_join(kw.test.all.eff.size, by="variable") %>% 
  filter(p.adj < 0.05)

## Number of variables with p.adj-value <0.05
length(combine.kw$variable)

## table summarizing effect size of the significant variables 
table(combine.kw$magnitude)

### Post-hoc analysis dunn's test, pairwise multiple comparisons to check which are significant in pairwise analysis.
dun.test.pos <- df_stat_melt %>% 
  group_by(variable) %>% 
  dunn_test(value ~ cluster_number, p.adjust.method = "BH", detailed = TRUE)
dun.test.pos <- dun.test.pos %>% 
  filter(p.adj<0.05) # %>% 
  #rename(cluster_number_1 = group1) %>% 
  #rename(cluster_number_2 = group2)
dun.test.pos

```

### D28
```{r}
## select the dataset
df_stat <- immunotype %>% 
  left_join(HI_df, by="subject_identifier") %>%
  drop_na(cluster_number) %>%
  dplyr::select(subject_identifier,cluster_number,GMT_T4) %>%
  droplevels()

## subject identifier and cluster number as.character
df_stat %<>% mutate_at(
  c("subject_identifier","cluster_number"),
  list(~as.character(.)))

## melt
df_stat_melt <- df_stat %>% 
  reshape2::melt(.)

### kruskal.test and Dunn's test
## With kruskal.test we identify overall which subsets are significant and then Dunn's test tells us which are different between pairs of clusters
kw.test.all <- df_stat_melt %>% 
  group_by(variable) %>% 
  kruskal_test(value ~ cluster_number) %>% 
  adjust_pvalue(method = "BH") 
kw.test.all.eff.size <- df_stat_melt %>% 
  group_by(variable) %>% 
  kruskal_effsize(value ~ cluster_number) 
combine.kw <- kw.test.all %>% 
  left_join(kw.test.all.eff.size, by="variable") %>% 
  filter(p.adj < 0.05)

## Number of variables with p.adj-value <0.05
length(combine.kw$variable)

## table summarizing effect size of the significant variables 
table(combine.kw$magnitude)

### Post-hoc analysis dunn's test, pairwise multiple comparisons to check which are significant in pairwise analysis.
dun.test.pos <- df_stat_melt %>% 
  group_by(variable) %>% 
  dunn_test(value ~ cluster_number, p.adjust.method = "BH", detailed = TRUE)
dun.test.pos <- dun.test.pos %>% 
  filter(p.adj<0.05) # %>% 
  #rename(cluster_number_1 = group1) %>% 
  #rename(cluster_number_2 = group2)
kw.test.all
dun.test.pos
```

### D28D0
```{r}
## select the dataset
df_stat <- immunotype %>% 
  left_join(HI_df, by="subject_identifier") %>%
  drop_na(cluster_number) %>%
  dplyr::select(subject_identifier,cluster_number,GMT_T4T0) %>%
  droplevels()

## subject identifier and cluster number as.character
df_stat %<>% mutate_at(
  c("subject_identifier","cluster_number"),
  list(~as.character(.)))

## melt
df_stat_melt <- df_stat %>% 
  reshape2::melt(.)

### kruskal.test and Dunn's test
## With kruskal.test we identify overall which subsets are significant and then Dunn's test tells us which are different between pairs of clusters
kw.test.all <- df_stat_melt %>% 
  group_by(variable) %>% 
  kruskal_test(value ~ cluster_number) %>% 
  adjust_pvalue(method = "BH") 
kw.test.all.eff.size <- df_stat_melt %>% 
  group_by(variable) %>% 
  kruskal_effsize(value ~ cluster_number) 
combine.kw <- kw.test.all %>% 
  left_join(kw.test.all.eff.size, by="variable") %>% 
  filter(p.adj < 0.05)

## Number of variables with p.adj-value <0.05
length(combine.kw$variable)

## table summarizing effect size of the significant variables 
table(combine.kw$magnitude)

### Post-hoc analysis dunn's test, pairwise multiple comparisons to check which are significant in pairwise analysis.
dun.test.pos <- df_stat_melt %>% 
  group_by(variable) %>% 
  dunn_test(value ~ cluster_number, p.adjust.method = "BH", detailed = TRUE)
dun.test.pos <- dun.test.pos %>% 
  filter(p.adj<0.05) # %>% 
  #rename(cluster_number_1 = group1) %>% 
  #rename(cluster_number_2 = group2)

kw.test.all
dun.test.pos

```

# Data organization, trucount baseline and post-vaccination
```{r}
## Trucounts
truc_all <- vital.truc_influ_additional 

truc_d0 <- vital.truc_influ_d0_additional %>%
  dplyr::select(-timepoint) %>%
  as.data.frame() 

truc_d2 <- vital.truc_influ_d2_additional %>%
  dplyr::select(-timepoint) %>%
  as.data.frame()  

truc_d7 <- vital.truc_influ_d7_additional %>%
  dplyr::select(-timepoint) %>%
  as.data.frame()

truc_d1d0 <- truc_d1d0_raw_p01 %>% 
  as.data.frame() %>%
  drop_na()

truc_d2d0 <- truc_d2d0_raw_p01 %>% 
  as.data.frame() %>%
  drop_na()

truc_d12d0 <- truc_d12d0_raw_p01 %>% 
  as.data.frame() %>%
  drop_na()

truc_d1d0_log2 <- df_kinetics_D1_D0_influ_log2fc_p01 %>% 
  as.data.frame() %>%
  drop_na()

truc_d2d0_log2 <- df_kinetics_D2_D0_influ_log2fc_p01 %>% 
  as.data.frame() %>%
  drop_na()

truc_d12d0_log2 <- df_kinetics_D12_D0_influ_log2fc_p01 %>% 
  as.data.frame() %>%
  drop_na()

truc_d7d0_log2 <- df_kinetics_D7_D0_influ_log2fc_p01 %>% 
  as.data.frame() %>%
  drop_na()

truc_d7d0 <- truc_d7d0_raw_p01 %>% 
  as.data.frame() %>%
  drop_na()

```

# Data organization for models
```{r}
## df_model dataset, HI and age,sex
df_model_1 <- as.data.frame(HI_df_subject) %>%
  dplyr::select(-c(GMT_T5,GMT_T5T0,strain))%>%
  dplyr::mutate(sex = dplyr::recode(sex, "1" = 1, "2" = 2),
                age_group = dplyr::recode(age_group, "25-49" = 1, "50-64"= 2, "65-98"= 3),
                T0_l3 = dplyr::recode(T0_l3,"<40" = 1, "40-80"= 2, ">80" = 3)) %>%
  mutate_at(c("sex","age_group","T0_l3"), list(~as.factor(.))) %>%
  drop_na(GMT_T4,GMT_T4T0) %>%
  droplevels()


## df_model dataset, HI and age,sex, CVI
df_model_2 <- as.data.frame(HI_df_subject_CVI) %>% 
  dplyr::select(-c(GMT_T5,GMT_T5T0,strain))%>%
  dplyr::mutate(sex = dplyr::recode(sex, "1" = 1, "2" = 2),
                age_group = dplyr::recode(age_group, "25-49" = 1, "50-64"= 2, "65-98"= 3),
                CMV = dplyr::recode(MIA_CMV_Seropositivity, "1" = 1, "2" = 2, "3" = 3),
                EBV = dplyr::recode(MIA_EBV_Seropositivity, "1" = 1, "2" = 2, "3" = 3),
                VZV = dplyr::recode(MIA_VZV_Seropositivity, "1" = 1, "2" = 2),
                T0_l3 = dplyr::recode(T0_l3,"<40" = 1, "40-80"= 2, ">80" = 3)) %>%
  mutate_at(c("sex","age_group","T0_l3","CMV","EBV","VZV"), list(~as.factor(.))) %>% 
  filter(CMV %in% c(1,2)) %>%
  drop_na(GMT_T4,GMT_T4T0) %>%
  relocate(where(is.numeric), .after = last_col()) %>%
  droplevels()
df_model_2$CMV <- factor(df_model_2$CMV, levels = c(2,1))

## df_model dataset, HI and age,sex, CVI, immunotype
df_model_3 <- as.data.frame(HI_df_subject_CVI) %>%
  dplyr::select(-c(GMT_T5,GMT_T5T0,strain))%>%
  left_join(immunotype, by="subject_identifier") %>%
  dplyr::mutate(sex = dplyr::recode(sex, "1" = 1, "2" = 2),
                age_group = dplyr::recode(age_group, "25-49" = 1, "50-64"= 2, "65-98"= 3),
                T0_l3 = dplyr::recode(T0_l3,"<40" = 1, "40-80"= 2, ">80" = 3)) %>%
  mutate_at(c("sex","cluster_number","age_group","T0_l3"), list(~as.factor(.))) %>%
  drop_na(GMT_T4,GMT_T4T0,cluster_number) %>%
  relocate(where(is.numeric), .after = last_col())%>%
  droplevels()

## df_model dataset, HI and age,sex, CVI and d0 truc
df_model_6 <- as.data.frame(HI_df_subject_CVI) %>% 
  dplyr::select(-c(GMT_T5,GMT_T5T0,strain))%>%
  left_join(vital.truc_influ_d0_additional, by="subject_identifier") %>%
  dplyr::select(-timepoint) %>%
  dplyr::mutate(sex = dplyr::recode(sex, "1" = 1, "2" = 2),
                age_group = dplyr::recode(age_group, "25-49" = 1, "50-64"= 2, "65-98"= 3),
                CMV = dplyr::recode(MIA_CMV_Seropositivity, "1" = 1, "2" = 2, "3" = 3),
                EBV = dplyr::recode(MIA_EBV_Seropositivity, "1" = 1, "2" = 2, "3" = 3),
                VZV = dplyr::recode(MIA_VZV_Seropositivity, "1" = 1, "2" = 2),
                T0_l3 = dplyr::recode(T0_l3,"<40" = 1, "40-80"= 2, ">80" = 3)) %>%
  mutate_at(c("sex","age_group","T0_l3","CMV","EBV","VZV"), list(~as.factor(.))) %>% 
  filter(CMV %in% c(1,2)) %>%
  relocate(where(is.numeric), .after = last_col())%>%
  drop_na(GMT_T4,GMT_T4T0,granulocytes_D0_per) %>%
  droplevels()

## df_model dataset, HI and age,sex, CVI and d2d0 truc
df_model_9 <- as.data.frame(HI_df_subject_CVI) %>% 
  dplyr::select(-c(GMT_T5,GMT_T5T0,strain))%>%
  left_join(truc_d12d0_log2, by="subject_identifier") %>%
  dplyr::mutate(sex = dplyr::recode(sex, "1" = 1, "2" = 2),
                age_group = dplyr::recode(age_group, "25-49" = 1, "50-64"= 2, "65-98"= 3),
                CMV = dplyr::recode(MIA_CMV_Seropositivity, "1" = 1, "2" = 2, "3" = 3),
                EBV = dplyr::recode(MIA_EBV_Seropositivity, "1" = 1, "2" = 2, "3" = 3),
                VZV = dplyr::recode(MIA_VZV_Seropositivity, "1" = 1, "2" = 2),
                T0_l3 = dplyr::recode(T0_l3,"<40" = 1, "40-80"= 2, ">80" = 3)) %>%
  mutate_at(c("sex","age_group","T0_l3","CMV","EBV","VZV"), list(~as.factor(.))) %>% 
  filter(CMV %in% c(1,2)) %>%
  relocate(where(is.numeric), .after = last_col())%>%
  drop_na(GMT_T4,GMT_T4T0,granulocytes_D12D0_per) %>%
  droplevels()

## df_model dataset, HI and age,sex, CVI and d7d0 truc
df_model_5 <- as.data.frame(HI_df_subject_CVI) %>% 
  dplyr::select(-c(GMT_T5,GMT_T5T0,strain))%>%
  left_join(truc_d7d0_log2, by="subject_identifier") %>%
  dplyr::mutate(sex = dplyr::recode(sex, "1" = 1, "2" = 2),
                age_group = dplyr::recode(age_group, "25-49" = 1, "50-64"= 2, "65-98"= 3),
                CMV = dplyr::recode(MIA_CMV_Seropositivity, "1" = 1, "2" = 2, "3" = 3),
                EBV = dplyr::recode(MIA_EBV_Seropositivity, "1" = 1, "2" = 2, "3" = 3),
                VZV = dplyr::recode(MIA_VZV_Seropositivity, "1" = 1, "2" = 2),
                T0_l3 = dplyr::recode(T0_l3,"<40" = 1, "40-80"= 2, ">80" = 3)) %>%
  mutate_at(c("sex","age_group","T0_l3","CMV","EBV","VZV"), list(~as.factor(.))) %>% 
  filter(CMV %in% c(1,2)) %>%
  relocate(where(is.numeric), .after = last_col())%>%
  drop_na(GMT_T4,GMT_T4T0,granulocytes_D7D0_per) %>%
  droplevels()
```



